---
title: "Bellabeat Capstone Project"
author: "Kaloyan Kirilov"
date: "`r Sys.Date()`"
output: html_document
---

# Bellabeat Capstone Project in R

## Introduction to the Project

The following represents an overview of the key parts of a case study with which I was tasked at the final stage of the Google Data Analytics Professional Certificate. This project analyses data of more than 30 FitBit users in the span of a month in order to gain insights that would inform the growth and marketing strategies of Bellabeat. 

### About the company

Bellabeat is a high-tech manufacturer of health-focused products for women, such as a wellness tracker, a wellness watch, a water bottle that tracks daily water intake, and an app that provides users with health data.

As the company is looking for more opportunities for growth, they have tasked me with analyzing smart device usage data in order to identify consumer trends and propose high-level recommendations to inform the Bellabeat's marketing strategy.

### Business task

Identify new growth opportunities in order to scale the business and become a larger player in the global smart device market.

I will do this by identifying consumer trends in wellness smart device usage and propose recommendations with the aim to exploit those trends.

### About the data

The [FitBit Fitness Tracker Data](https://www.kaggle.com/datasets/arashnic/fitbit) is a publicly available data set generated by 30 respondents to a distributed survey via Amazon Mechanical Turk between 12 March 2016 and 12 May 2016. The dataset is published on Kaggle under a CC0: Public Domain licence with no citation guidelines.

After having determined the questions we will answer through this data, it is time to proceed to the data preparation stage.

## Data preparation

The data has been downloaded and stored into a local hard disk. No additional security measures seem necessary as the data is openly available for download. The data itself is anonymized, which adds a layer of protection for the subjects whose data I will be working with.

The data is not current, as it was gathered in 2016, while the analysis is performed in 2023. In addition, its reliability is questionable as it reflects consumer usage of FitBit devices, yet it is collected via an Amazon Mechanical Turk survey and uploaded on Kaggle. In this context, it is impossible to properly verify the data's integrity.

Further limitations linked to the dataset are reified by the small sample size - 33 subjects have shared their data for the majority of the aspects being tracked, such as a general daily overview of the subjects' activities, as well as hourly data for their intensitieis, steps, and calories.

Meanwhile, there were only 14 subjects that have shared their heart rate data, 24 subjects sharing their sleep data, and 8 subjects having shared weight data. All the data is also limited in time, as it tracks only a one month time period (April - May 2016).

These limitations pose serious concerns over the integrity and reliability of the data and the conclusions that can be gathered thereof, yet do not take away from the main purpose of this exercise - namely, to apply and showcase some of the skills learned during the Google Data Analytics Professional Certificate.

Next, I will move on to the data processing stage of the process.

## Data processing

### Loading packages

```{r loading packages}
library(tidyverse)
library(ggpubr)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(chron)
library(plyr)
library(anytime)
library(data.table)
library(hms)
library(gridExtra)
```

### Importing the relevant data sets

While there are 18 data sets in the database, including a daily activity data set, daily, hourly, and minutely data sets for calories, intensities, and steps, as well as daily sleep data set, weight data set, and secondly heartrate data set.

I will exclude some data sets due to the small sample size, such as the weigh data set, and data sets that comprise of data already included in a more general data set, such as with the daily data for calories included in activity.

```{r}
heartrate_seconds <- read.csv("E:/Work/Data Analysis/Google Data Analytics Certificate/Course 8 - Capstone project/Bellabeat Project/Fitbit Dataset/heartrate_seconds_merged.csv")
activity <- read.csv("E:/Work/Data Analysis/Google Data Analytics Certificate/Course 8 - Capstone project/Bellabeat Project/Fitbit Dataset/dailyActivity_merged.csv")
calories <- read.csv("E:/Work/Data Analysis/Google Data Analytics Certificate/Course 8 - Capstone project/Bellabeat Project/Fitbit Dataset/hourlyCalories_merged.csv")
intensities <- read.csv("E:/Work/Data Analysis/Google Data Analytics Certificate/Course 8 - Capstone project/Bellabeat Project/Fitbit Dataset/hourlyIntensities_merged.csv")
steps <- read.csv("E:/Work/Data Analysis/Google Data Analytics Certificate/Course 8 - Capstone project/Bellabeat Project/Fitbit Dataset/hourlySteps_merged.csv")
sleep <- read.csv("E:/Work/Data Analysis/Google Data Analytics Certificate/Course 8 - Capstone project/Bellabeat Project/Fitbit Dataset/sleepDay_merged.csv")
```

### Cleaning the data

#### Checking for duplicates 

```{r}
sum(duplicated(heartrate_seconds))
sum(duplicated(activity))
sum(duplicated(calories))
sum(duplicated(intensities))
sum(duplicated(steps))
sum(duplicated(sleep))
```

#### Removing duplicates

```{r}
sleep <- distinct(sleep)
sum(duplicated(sleep))
```

#### Exploring the data 

```{r}
glimpse(heartrate_seconds)
glimpse(activity)
glimpse(calories)
glimpse(intensities)
glimpse(sleep)
```

While also using the str() and head() functions, I observed that the 'date' and 'time' columns in each table are in the 'chr' data type, which will cause issues with the analysis down the road. In addition, some data sets have the 'time' varibale, while others do not. Further, the name of the variables is not consistent along all the data sets. Thus, the data requries further cleaning and transformation before being ready to be merged.

#### Cleaning the date and time irregularities
#### Renaming columns
#### Excluding irrelevant columns from some data sets

```{r}
# For 'activity'
activity$ActivityDate=as.POSIXct(activity$ActivityDate, format="%m/%d/%Y", tz="UTC")
colnames(activity)[colnames(activity) == "ActivityDate"] = "Date"

# For 'calories'
calories$ActivityHour=as.POSIXct(calories$ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz="UTC")
calories$Date <- as.Date(calories$ActivityHour, format = "%m/%d/%y", tz="UTC")
calories$Time <- times(strftime(calories$ActivityHour, "%H:%M:%S", tz="UTC"))
calories <- calories[, c(1,4,5,3)]

# For 'intensities'
intensities$ActivityHour=as.POSIXct(intensities$ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz="UTC")
intensities$Date <- as.Date(intensities$ActivityHour, format = "%m/%d/%y", tz="UTC")
intensities$Time <- times(strftime(intensities$ActivityHour, "%H:%M:%S", tz="UTC"))
intensities <- intensities[, c(1, 5, 6, 3, 4)]

# For 'sleep'
sleep$SleepDay=as.POSIXct(sleep$SleepDay, format="%m/%d/%Y %I:%M:%S %p", tz="UTC")
colnames(sleep)[colnames(sleep) == "SleepDay"] = "Date"
sleep$Date <- as.Date(sleep$Date, format = "%Y-%m-%d", tz="UTC")

# For 'steps'
steps$ActivityHour=as.POSIXct(steps$ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz="UTC")
steps$Date <- as.Date(steps$ActivityHour, format = "%m/%d/%y", tz="UTC")
steps$Time <- times(strftime(steps$ActivityHour, "%H:%M:%S", tz="UTC"))
steps <- steps[, c(1, 4, 5, 3)]
```

The heartrate data set would require special attention as I also intend to aggregate the data in average hourly heartrate in order to merge it with the other hourly data sets.

```{r}
# For heartrate_seconds
heartrate_seconds$Datetime=as.POSIXct(heartrate_seconds$Time, format="%m/%d/%Y %I:%M:%S %p", tz="UTC")
heartrate_seconds <- heartrate_seconds[, c(1, 4, 3)]
heartrate_seconds$Datetime2 <- droplevels(cut(heartrate_seconds$Datetime, breaks="hour"))
heartrate_seconds$Datetime2 = as.POSIXct(heartrate_seconds$Datetime2, format="%Y-%m-%d %H:%M:%S", tz='UTC')
heartrate_hourly <- heartrate_seconds %>%
    group_by(Id, Datetime2) %>%
    aggregate(Value ~ Id + Datetime2, FUN=mean)
heartrate_hourly$Date <- as.Date(anytime(as.factor(heartrate_hourly$Datetime2), tz='UTC'), format = '%Y-%m-%d', tz='CET')
heartrate_hourly$Time <- times(strftime(heartrate_hourly$Datetime2, format = '%H:%M:%S', tz='UTC'))
heartrate_hourly <- heartrate_hourly[, c(1, 3, 5, 4)]
```

## Analyzing the data

The analysis of the data requries for it to be formatted, organized, sorted, and transformed.

I have already renamed and formated some of the data sets in the previous step.

In addition, I observed that there are records where (a) the expended daily calories are 0, (b) the total amount of minutes recorded are less than 120, or (c) the daily expended calories are less than 200, which is impossible for a living person. This could be due to, among others, detaching the tracking device from one's body, battery shortage, or faulty device. In any case, those 9 data points are outliers and I will filter them out for the sake of a cleaner data set.

```{r}
activity <- filter(activity, VeryActiveMinutes+FairlyActiveMinutes+LightlyActiveMinutes+SedentaryMinutes > 120 & Calories > 200)
```

Further, I will proceed to joining several of the data sets together, which will make working with them easier. I will combine all the data sets that are based on temporal characteristcs. Thus, I will merge the general daily activity data set with the daily sleep data set. Then, I will merge hourly data sets for calories, intensities, steps, and heartrate. As there are fewer observations for the heartrate data set than for the other three, I will create two separate tables - one containing the three horly data sets except heartrate, and one extended table including heartrate.

_Note that the heartrate data set has already been transformed from secondly to hourly in the previous stage so as to allow joining it with the rest._

### Merging data sets

```{r}
merged_activity_sleep <- inner_join(sleep, activity, by=c('Id', 'Date'))
merged_hourly <- join_all(list(calories, intensities, steps), by=c('Id', 'Date', 'Time'))
merged_hourly_extended <- inner_join(merged_hourly, heartrate_hourly, by=c('Id', 'Date', 'Time'))
colnames(merged_hourly_extended)[colnames(merged_hourly_extended) == "Value"] = "AverageHeartrate" 
```

```{r}
summary(merged_activity_sleep)
```
```{r}
summary(merged_hourly)
```

### Additional sorting and transformation

The data needs to be further sorted and transformed in order to be able to identify patterns, such as:

* days of week in which users take the most steps
* days of the week in which users spent the most amount of time sleeping
* days of the week in which users are most active (in terms of recorded 'intensities')
* days of the week in which users exhaust the most amount of calories
* hours in which users are the most active (in terms of recorded 'intensities')

```{r}
# This code links each data point with the day of the week
# assosicated with its date. Then, it gives an order to the list
# based on the order of the days of the week. Finally, it proceeds
# to group all the data points by the associated day of the week
# and aggregating the data to find the mean for the total steps
# and total time spent in bed associated which each day of the week.

weekday_steps_sleep <- merged_activity_sleep %>% 
  mutate(weekday = weekdays(Date))

weekday_steps_sleep$weekday <- ordered(weekday_steps_sleep$weekday, 
                                       levels=c("Monday", "Tuesday", "Wednesday",
                                                "Thursday", "Friday", "Saturday",
                                                "Sunday"))

weekday_steps_sleep <- weekday_steps_sleep %>% 
  dplyr::group_by(weekday) %>% 
  dplyr::summarise(mean_steps = mean(TotalSteps), mean_sleep = mean(TotalMinutesAsleep))

head(weekday_steps_sleep)
```


```{r}
# Similar approach to aggregate the hourly data for Calories and TotalIntensity
# into daily data and see how much times users are recorded to be performing a more 
# intense activity and the calories they exhaust during the week.

weekday_intensity_calories <-  merged_hourly %>% 
  group_by(Id, Date) %>% 
  aggregate(cbind(Calories, TotalIntensity) ~ Id + Date, FUN=sum)

weekday_intensity_calories <- weekday_intensity_calories %>% 
  mutate(weekday = weekdays(Date))

weekday_intensity_calories$weekday <- ordered(weekday_intensity_calories$weekday, 
                                              levels=c("Monday", "Tuesday", "Wednesday",
                                                       "Thursday", "Friday", "Saturday",
                                                       "Sunday"))

weekday_intensity_calories <- weekday_intensity_calories %>% 
  dplyr::group_by(weekday) %>% 
  dplyr::summarise(mean_intensities = mean(TotalIntensity), mean_calories = mean(Calories))

head(weekday_intensity_calories)

```

```{r}
# Creating a data frame where we can group the data by every hour
# and find the mean of intensities for each hour.

track_int <- intensities %>% 
  select(Time, TotalIntensity) %>% 
  dplyr::group_by(Time) %>% 
  dplyr::summarise(mean_intensity = mean(TotalIntensity, na.rm = TRUE))

# Thereafter, we need to format the time appropriately:

track_int$Time2 <- as.character(track_int$Time)
track_int$Time2 <- as.hms(track_int$Time2)
```

### Summary of the analysis

The analysis that will be visualized below confirmed some logical hypotheses, such as the existence of a positive correlation between, among others:

* the number of intensities subjects had during the day and their heartrate
* the amount of steps people made and the calories they exhausted

Interestingly, I observed that there is a significant negative correlation between the amount of minutes subjects spent in a sedentary mode, presumably seated, laying, or overall relaxed and immobile, and the time they slept throughout the night.

Overall, tracking device users were most active between 5 and 8 pm, noting that there is also a measurable increase in activity around 12 pm.

Subjects made the most amount of steps on Saturday, with Monday and Tuesday coming in short second and third, respecitvely. Likewise, users are performing the most amount of intense activities on Saturday, Tuesday, and Monday.

Conversely, users are least active on Sunday both in terms of steps taken, as well as the amount of intensities recorded.

This can be linked to the observation that users tend to sleep the most on Sunday, and, with the exception of Wednesday, sleep less during the week.

Calories-wise, similar patterns are present. The days in which users exhaust the most amount of calories are Saturday, Tuesday, and Monday, while Sunday and Thursday are on the opposite end.

These trends represent the key insights that would allow us to answer the questions the business is asking and inform the future growth strategy of the company.

## Sharing the data

### General observations

As we can see below, users have burnt more calories when they have taken more steps during the day. Likewise, their heart rate increases with the amount of intense activities they perform during the day.

This does not mean that people who are more active tend to have a higher resting heart rate. While this could be further investigated, for which a bigger sample size would be beneficial, the scientific community suggest that people who are more active tend to have lower resting heart rates, as shown in [this](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6306777/) study.

```{r echo = FALSE}
# Relationship between the amount of steps and the calories burnt.

ggplot(activity, aes(TotalSteps, Calories))+
  geom_point()+
  geom_smooth()+
  labs(x = "Total steps", y = "Calories", title="Total steps vs. calories")
```

```{r echo = FALSE}
# Relationship between the amount of intensities and the average heartrate of the users

ggplot(merged_hourly_extended, aes(AverageIntensity, AverageHeartrate))+
  geom_point() +
  geom_smooth()+
  facet_wrap(merged_hourly_extended$Id) +
  labs(x = 'Mean hourly level of intensity', 
       y = 'Mean hourly heart rate', 
       title = 'Relationship between measured heart rate and level of intensity', 
       subtitle = 'Hourly mean data concerning the level of intensity and associated heart rate for 14 subjects tracked between 12/04/2016 - 12/05/2016', 
       caption = 'Pulic dataset: https://www.kaggle.com/datasets/arashnic/fitbit ')

```

### Sleep

When it comes to sleep, I was unable to detect a significant relationship between the amount spent asleep and the amount of minutes subjects have been very, fairly, or ligthly active.

Meanwhile, there is a significant negative correlation between the amount of minues people have stayed sitted or more generally inactive and the amount of time they have spent sleeping.

```{r echo = FALSE}
ggplot(merged_activity_sleep, aes(SedentaryMinutes, TotalMinutesAsleep))+
  geom_point(color="blue")+
  geom_smooth(color="red", method = "loess", formula = "y~x")+
  labs(x = "Sedentary minutes", 
       y = "Sleep time ", 
       title = "Sleep time vs. Sedentary miutes", 
       caption = 'Pulic dataset: https://www.kaggle.com/datasets/arashnic/fitbit')
```

Thus, this data suggests that people sleep less when they have spent more minutes in a sedentary mode during the day, which leads to the conclusion that being more active improves sleep in terms of quantity.

It would be great to dive further into this topic and acquire data that would allow us to measure the quality of sleep and research potential positive correlations between more physical activity and better sleep in terms of quality and/or quantity.

### When are users most active

In this section, we focus on the trends concerning users most active days and hours.

```{r echo = FALSE}
# Illustrating two graphs next to each other - one showing 'Daily steps per weekday'
# and the other focusing on 'Minutes asleep per weekday'

ggarrange(
  ggplot(weekday_steps_sleep, aes(weekday, mean_steps)) +
    geom_col(fill = "darkblue") +
    geom_hline(yintercept = 7500) +
    labs(x = "Day of the week", y = "Mean - Total Steps", title = "Daily steps per weekday") +
    theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 0.4)),
  ggplot(weekday_steps_sleep, aes(weekday, mean_sleep)) +
    geom_col(fill = "darkcyan") +
    geom_hline(yintercept = 480) +
    labs(x = "Day of the week", y = "Mean - Minutes Asleep",title = "Minutes asleep per weekday") +
    theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 0.4))
)

```

The graphs above ilustrate several interesting trends.

On the left, we can see the amount of steps users have taken on average during each day of the weekday during the duration of the study. We obsere that Saturday is - by far - the most active day, followed by Monday and Tuesday, whereas Sunday is the day where users have made the least amount of steps.

This insight can be linked with what is depicted on the right, namely that users tend to sleep the most on Sundays. More sleep correlates with less steps.

Note that Sunday is the only day where users have not met the quota, albeit artificial, of steps that make for amount to be considered 'somwhat active' as per this categorization made by 1000 Steps.

```{r echo = FALSE}
# Graphs illustrating the 'Total intensities per weekday'
# and the 'Calories exhausted per weekday'

ggarrange(
  ggplot(weekday_intensity_calories, aes(weekday, mean_intensities)) +
    geom_col(fill = "darkblue") +
    labs(x = "Day of the week", y = "Mean - Total Intensities", title = "Total intensities per weekday") +
    theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 0.4)),
  ggplot(weekday_intensity_calories, aes(weekday, mean_calories)) +
    geom_col(fill = "darkcyan") +
    labs(x = "Day of the week", y = "Mean - Calories",title = "Calories exhausted per weekday") +
    theme(axis.text.x = element_text(angle = 45,vjust = 0.5, hjust = 0.4))
)
```

These two graphs confirm the story told by the previous visualization.

Users tend to be most active on Saturday, Tuesday, and Monday and least active on Sunday and during the middle of the work week.

Let's see in which hours they are the most and least active.

```{r echo = FALSE}
# A histogram visualizing users' most  and least active hours

ggplot(track_int, aes(Time2, mean_intensity))+
  geom_histogram(stat = "identity", fill = "darkblue")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))+
  labs(x = "Time of day",
       y= "Mean - Intensities", 
       title = "At what time are users most active throughout the day?",
       caption ="Pulic dataset: https://www.kaggle.com/datasets/arashnic/fitbit")
```

Users are most active between 5 pm and 8 pm, with a noteworthy spike in activitiy midday, around 12 pm. Conversely, activity drops right after 9 and plummets to near 0 between 12 am and 4 am, when people generally tend to sleep.

With this, we have a fairly comprehensive overview of the hours and days users tend to be most and least active. These trends can be leveraged in many ways by Bellabeat, including in its communcation with its users, by giving them a nudge to plan their acitivities around those hours and notify them when those hours and days approach.

Let's move to the recommendations stage of this presentation in order to dive deeper into this.

## Act - recommendations

In conclusion, we were able to identify some useful trends regarding users' tracking device usage behavior, as well as some insights related to the impact movement can have on sleep, calories burnt, and the heart rate of users.

These insights and trends can be applied to Bellabeat users in several ways, with the list being non-exhaustive:

Bellabeat can otpimize its communication with users throught the company's app so as to give more relevant notifications and be more useful for its customers by taking into account when users are most and least active

The app can give more elaborate general advice on the topics of exercise, sleep, and heart rate management to its users and, thus, increase its user-friendliness and overall usefulness for customers

Along the lines of the previous point, Bellabeat can give better personalized advice to its customers based on theri individual goals and preferences *(i.e., prompt a customer who wants to sleep more to increase its active minutes, even if by the slightest, as this will increase their time spent asleep) *

In terms of marketing strategy, Bellabeat can tweak their customer acquisition operations to be more aggressive on Saturdays, Mondays, and Tuesdays, and significantly lowering their efforts on Sundays and in the middle of the week

In addition to those concrete recommendations focused on Bellabeats attitude towards existing and potential customers, I would also recommend the company to:

* expand on efforts to gather data that is more representative, with bigger sample size, and more current, as this will significantly improve the data analysis process
* expand on the data being gathered itself, by including more variables that would allow for deeper analysis

### Credits

As this case study is part of the Google Data Analytics Professional Certificate, there are many individuals who have already had their own take on the project and have shared it publicly. Some of my work has been inspired by some of the projects I have seen both Kaggle and beyond and would like to thank the community for their work. In particular, I want to thank [Anastasiia Chebotina](https://www.kaggle.com/code/chebotinaa/bellabeat-case-study-with-r) and [Macarena Lacasa](https://www.kaggle.com/code/macarenalacasa/capstone-case-study-bellabeat) for their insightful and interesting work.